---
title: "Blueberry Differential Expression Analysis with EdgeR"
author: "Scott Teresi"
affiliation: "Edger Lab"
abstract:
thanks:
keywords:
output:
  pdf_document:
    highlight: default
    citation_package:
    keep_tex: false
    fig_caption: true
    latex_engine: pdflatex
fontsize: 12pt
urlcolor: blue
geometry: margin=1in
bibliography:
biblio-style:
header-includes:
- \usepackage{indentfirst}
- \usepackage{graphicx}
- \usepackage{geometry}
- \usepackage{subfigure}
- \usepackage{amsmath}
- \usepackage{listings}
- \usepackage{tikz}
- \usetikzlibrary{matrix}
---

\vspace{0.25in}

# Purpose:
Produce a matrix of differentially expressed genes for the blueberry RNA data, a smear plot, and a summary table. I will be using the package [edgeR](https://www.bioconductor.org/packages/release/bioc/vignettes/edgeR/inst/doc/edgeRUsersGuide.pdf) for this task.

# Accessibility and Help:
The following guide, source code, and other components of the blueberry expression analysis pipeline can be found at [Blueberry_RNA_Seq_Expression_Analysis](https://github.com/EdgerLab/Blueberry_RNA_Seq_Expression_Analysis) Github page. This page includes general information, the keys of the samples, and the list of comparisons desired.

# Installation and Loading of Library:
Using this [link](https://bioconductor.org/packages/release/bioc/html/edgeR.html) as reference, please install **edgeR**. I have set the chunk evaluation clause to FALSE, as you only need to install once and should do it manually.
```{r eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("edgeR")
```

We also need [tidyverse](https://www.tidyverse.org/) which includes [dplyr](https://dplyr.tidyverse.org/) but just to be safe we will install both with:
```{r eval=FALSE}
install.packages("tidyverse")
install.packages("dplyr")
```

And then we will load the libraries with:
```{r message=FALSE}
library(edgeR)
library(tidyverse)
library(dplyr)
```

# Loading the Data Into R (Part 1):
Here we will load all of our data from HTSeq into R. There are 3 files. Two of which, the 802 and 724 data sets can be lumped together as they have similar samples. The 809 data set belongs on its own. Below I import the data, perform the appropriate merges and reorient the data. I also remove rows that are completely 0s, because they are uninformative rows (genes) and will cause statistical issues.

```{r Load_Data}
# Import data
load_data = function(data_dir){
	Blueberry = read.csv(data_dir, sep='\t', header=TRUE)
	#Blueberry = read.csv('/home/scott/Documents/Uni/Research/Projects/Blueberry_RNA_Seq_Expression_Analysis/Input/Count/Collate/SingleHaplotype_Counts_Blueberry.tsv',
	       #sep = '\t',
			   #header = TRUE)

	# Remove the extraneous 5 rows at the bottom, these extraneous are supplementary info from HTSeq
	Blueberry = head(Blueberry, n = -5)

	# Set the index
	rownames(Blueberry) = Blueberry$Gene

	# Get rid of gene name column because it is now the index
	Blueberry = subset(Blueberry, select = -c(Gene))

	# Remove the rows that are completely 0s, uninformative rows
	Blueberry = Blueberry[apply(Blueberry, 1, function(x) {
	  !all(x == 0)}), ] 
return(Blueberry)
}
```

# Loading the Data Into R (Part 2):
Here we will filter the data and add the appropriate "metadata" so that we can easily recognie each sample. I write these as a function so that we can easily utilize them later inside the individual sample comparison chunks.

```{r Clean_Data, eval=TRUE}
clean_data = function(Blueberry){
	Metadata_Blueberry = Blueberry
	# Add two empty rows at the end of the data frame to be filled with the
	# experimental factors that we later plug in.
	Metadata_Blueberry[(nrow(Metadata_Blueberry) + 1):(nrow(Metadata_Blueberry) + 2), ] = NA

	# Loop through the columns and assign experimental factors based on the
	# sample names, filling the last two rows
	#columns = colnames(Metadata_Blueberry)
	dra_front_str = 'Dra_'
	lib_front_str = 'Lib_'
	stop_control_str = 'dpo_c'
	stop_treatment_str = 'dpo_t'
	 
	for (i in 1:ncol(Metadata_Blueberry)) {
	  for(xi in 1:7) {
	    xi = toString(xi)
			dra_full_C = paste(dra_front_str, xi, stop_control_str, sep = '')
			dra_full_T = paste(dra_front_str, xi, stop_treatment_str, sep = '')
			lib_full_C = paste(lib_front_str, xi, stop_control_str, sep = '')
			lib_full_T = paste(lib_front_str, xi, stop_treatment_str, sep = '')
			dra_name = paste('DRA_', xi, sep = '')
			lib_name = paste('LIB_', xi, sep = '')

			if (grepl(dra_full_C, colnames(Metadata_Blueberry)[i])) {	
			      Metadata_Blueberry[(nrow(Metadata_Blueberry) - 1),i] <- dra_name
			      Metadata_Blueberry[(nrow(Metadata_Blueberry)), i] <- "Control"

	      } else if (grepl(dra_full_T, colnames(Metadata_Blueberry)[i])) {
			Metadata_Blueberry[(nrow(Metadata_Blueberry) - 1),i] <- dra_name
			Metadata_Blueberry[(nrow(Metadata_Blueberry)), i] <- "Treatment"  

	      } else if (grepl(lib_full_C, colnames(Metadata_Blueberry)[i])) {
			Metadata_Blueberry[(nrow(Metadata_Blueberry) - 1),i] <- lib_name
			Metadata_Blueberry[(nrow(Metadata_Blueberry)), i] <- "Control"  

	      } else if (grepl(lib_full_T, colnames(Metadata_Blueberry)[i])) {
			Metadata_Blueberry[(nrow(Metadata_Blueberry) - 1),i] <- lib_name
			Metadata_Blueberry[(nrow(Metadata_Blueberry)), i] <- "Treatment"  
	      } 
	  }
	# Update rows with "metadata" on each sample's identity
	  row.names(Metadata_Blueberry)[(nrow(Metadata_Blueberry) - 1) : (nrow(Metadata_Blueberry))] = c("Identifier", "Experiment_Treatment")
	}
return(Metadata_Blueberry)
}
```

```{r CompareEm, eval = TRUE}
EdgeR_Func = function(Counts, G1, G2, xi, data_input_type, test_type, treatment_groupings_same=FALSE, comparison_order='Simple') {
	
    
	# Switch to output folder
  if (data_input_type == 'Single') {
      if (test_type == 'fdr') {
    	  setwd('/home/scott/Documents/Uni/Research/Projects/Blueberry_Data/Diff_Ex/EdgeR_Output/Single_Hap/FDR/')
    	}
    	else if (test_type == 'bonferroni') {
    	  setwd('/home/scott/Documents/Uni/Research/Projects/Blueberry_Data/Diff_Ex/EdgeR_Output/Single_Hap/Bonferroni/')
    	}
    
  } else if (data_input_type == 'All') {
      if (test_type == 'fdr') {
    	  setwd('/home/scott/Documents/Uni/Research/Projects/Blueberry_Data/Diff_Ex/EdgeR_Output/All_Hap/FDR/')
    	} else if (test_type == 'bonferroni') {
    	  setwd('/home/scott/Documents/Uni/Research/Projects/Blueberry_Data/Diff_Ex/EdgeR_Output/All_Hap/Bonferroni/')
    	}
  }
	  
	rownames(Counts) <- Counts$Row.names
	Counts = subset(Counts, select = -c(Row.names))
	if (isTRUE(treatment_groupings_same)) {
	  my_grouping = Counts['Identifier',] 
	} else {
	my_grouping = Counts['Experiment_Treatment',] # Make grouping for treatment groups.
	}
	Counts = Counts[!(rownames(Counts) %in% 'Identifier'),]
	Counts = Counts[!(rownames(Counts) %in% 'Experiment_Treatment'),]

	# Convert to matrix
	Counts = as.matrix.data.frame(Counts)
	x = rownames(Counts)
	# We need the gene names for later
	Gene_Row_Key = data.frame("Num"=rownames(as.data.frame(x)),"Gene"=x)
	rm(x)
	Counts = apply(Counts, 2, as.numeric)

	D = DGEList(Counts, group = unlist(my_grouping))
	D = calcNormFactors(D)
	D_Samples = D$samples
	D = estimateCommonDisp(D)
	D = estimateTagwiseDisp(D)
	if (comparison_order == 'Simple'){
	  Fish_Exact = exactTest(D, pair=c('Control', 'Treatment'))
	}
	else if (comparison_order == 'Complex'){
	  Fish_Exact = exactTest(D)
	}
	topTags = topTags(Fish_Exact)

	# P = 0.05
	if (test_type == 'fdr') {
	  simplified_DGE = decideTestsDGE(Fish_Exact, p=0.05, adjust='fdr')
	}
  else if (test_type == 'bonferroni') {
    simplified_DGE = decideTestsDGE(Fish_Exact, p=0.05, adjust='bonferroni')
	}


	# Write summary table
	write.table(summary(simplified_DGE), file = paste(G1, xi, '_vs_', G2, xi, '_Summary.txt', sep = ''), quote = F, sep = '\t')

	# Write direction of differential expression table
	simplified_DGE_frame = data.frame(simplified_DGE)
	colnames(simplified_DGE_frame) = 'Direction_Differentially_Regulated'
	row.names(simplified_DGE_frame) = Gene_Row_Key$Gene
	Direction = as_tibble(rownames_to_column(simplified_DGE_frame, var = 'Gene_Name'))
	write_tsv(Direction, paste(G1, xi, '_vs_', G2, xi, '_Direction.tsv', sep = ''))
}
```



# Compare C vs T
```{r Start_Comparisons, eval=TRUE}
c_obj_str = function(my_obj) {
  deparse(substitute(my_obj))
}

run_comparisons = function(data_input_type) {
	dra_front_str = 'Dra_'
	lib_front_str = 'Lib_'
	stop_control_str = 'dpo_c'
	stop_treatment_str = 'dpo_t'
	 
	for(xi in 1:7) {
	  xi = toString(xi)
		dra_full_C = paste(dra_front_str, xi, stop_control_str, sep = '')
		dra_full_T = paste(dra_front_str, xi, stop_treatment_str, sep = '')
		lib_full_C = paste(lib_front_str, xi, stop_control_str, sep = '')
		lib_full_T = paste(lib_front_str, xi, stop_treatment_str, sep = '')
		dra_name = paste('DRA_', xi, sep = '')
		lib_name = paste('LIB_', xi, sep = '')
		
	  DRA_C =  select(Metadata_Blueberry, matches(dra_full_C))
	  DRA_T =  select(Metadata_Blueberry, matches(dra_full_T))
	  LIB_C =  select(Metadata_Blueberry, matches(lib_full_C))
	  LIB_T =  select(Metadata_Blueberry, matches(lib_full_T))
	 
	  # DRA_C vs DRA_T 
	  #Loop this
	  Counts = merge(DRA_C, DRA_T, by = 'row.names')
	  G1 = c_obj_str(DRA_C)
	  G2 = c_obj_str(DRA_T)
	  EdgeR_Func(Counts, G1, G2, xi, data_input_type, 'bonferroni', comparison_order='Simple')
	  EdgeR_Func(Counts, G1, G2, xi, data_input_type, 'fdr', comparison_order='Simple')
	  
	  Counts = merge(LIB_C, LIB_T, by = 'row.names')
	  G1 = c_obj_str(LIB_C)
	  G2 = c_obj_str(LIB_T)
	  EdgeR_Func(Counts, G1, G2, xi, data_input_type, 'bonferroni', comparison_order='Simple')
	  EdgeR_Func(Counts, G1, G2, xi, data_input_type, 'fdr', comparison_order='Simple')
	  
	  Counts = merge(DRA_T, LIB_T, by = 'row.names')
	  G1 = c_obj_str(DRA_T)
	  G2 = c_obj_str(LIB_T)
	  EdgeR_Func(Counts, G1, G2, xi, data_input_type, 'bonferroni', treatment_groupings_same = TRUE, comparison_order='Complex')
	  EdgeR_Func(Counts, G1, G2, xi, data_input_type, 'fdr', treatment_groupings_same = TRUE, comparison_order='Complex')
	}
}
```

```{r execute_all, eval=TRUE}
single_hap = '/home/scott/Documents/Uni/Research/Projects/Blueberry_RNA_Seq_Expression_Analysis/Input/Count/Collate/SingleHaplotype_Counts_Blueberry.tsv'
Blueberry = load_data(single_hap)
Metadata_Blueberry = clean_data(Blueberry)
run_comparisons('Single')


all_hap = '/home/scott/Documents/Uni/Research/Projects/Blueberry_RNA_Seq_Expression_Analysis/Input/Count/Collate/AllCounts_Blueberry.tsv'
Blueberry = load_data(all_hap)
Metadata_Blueberry = clean_data(Blueberry)
run_comparisons('All')
```